---
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Methods description for Aloes paper

This document describes methods used in the paper **'Aloes Horn of Africa'**. Each heading relates to sections in the manuscript. See [Aloes_Horn_Diversity](https://github.com/stevenpbachman/Aloes_Horn_Diversity) for raw data and related documents.


### Libraries

```{r, message=FALSE}
library(dplyr)
library(tidyr)
library(readr)
library(ggplot2)
library(rgdal)
library(maptools)
library(sp)
library(rgeos)
library(sf)
library(rredlist)
```

### 1. Histogram of specimen collections 

*Aloe occurrence data*

Specimen data were derived from a database of collections from herbaria including ETH, K, EA, UPS, PRE, and some from FT and other institutions. The taxonomic treatments were verified by one of us (Demissew).
Occurrence data were submitted to the IUCN Red List as supporting information for the global Red List assessments. 

Occurrence data were queried from the [IUCN Red List](https://www.iucnredlist.org/) using the Advanced query and the following filters:

Taxonomy: Genus = Aloe
Land Regions: Country = Djibouti, Eritrea, Ethiopia, Somalia, South Sudan, Sudan 

```{r echo=TRUE, message = FALSE, warning=FALSE}
# import specimens
specimens_all<- read_csv("01_data/Aloe_points_csv.csv")

# add extra range data for two species that didn't have point data
specimens_extra = read_csv("01_data/Aloe_points_extra.csv")

# merge to get full dataset
all_occs = rbind(specimens_all,specimens_extra)

# find unique records (species) and count number of occurrences per species
spec_count = all_occs %>% group_by(binomial) %>% count()

# plot as histogram and show mean
ggplot(spec_count, aes(x = n)) + 
geom_histogram(color = 'darkgray', fill = 'white', binwidth = 2) +
labs(x = 'Number of occurrence records', y = 'Count of species') +
geom_vline(aes(xintercept = mean(spec_count$n, na.rm = TRUE), show.legend = FALSE)) +
scale_x_continuous(breaks = c(10, 20, 30, 40, 50, 60))

# save down the plot
path = getwd()
ggsave(paste0(path,"/03_figs/fig_1.png"), width = 10)

# calculate the mean number of occurrences per species
spec_mean = mean(spec_count$n)

```

### 2. Study area – show all points, and show study region countries where species are from.

Done in arcmap - explain

### 3. Aloe Red List assessment data - summarise categories and threat data:

Red List assessment data were queried from the [IUCN Red List](https://www.iucnredlist.org/) using the Advanced query and the following filters:

Taxonomy: Genus = Aloe
Land Regions: Country = Djibouti, Eritrea, Ethiopia, Somalia, South Sudan, Sudan 

The following tables were downloaded:
assessments: Species ID and Red List category
taxonomy: Species ID and taxonomic information (linked by species ID)

```{r echo=TRUE, message = FALSE, warning=FALSE}
# import data
assessments = read_csv("01_data/IUCN_downloads/assessments.csv")
taxonomy = read_csv("01_data/IUCN_downloads/taxonomy.csv")

# get single file with names and ratings
sp_ratings = merge(assessments, taxonomy, by = "internalTaxonId")

# update points with Red List Rating
colnames(all_occs)[which(names(all_occs) == "id_no")] = "internalTaxonId"
all_occs_rating = merge(all_occs, sp_ratings, by = "internalTaxonId")
```

*Summary of final Red List assessment ratings*

Replicate method to calculate table of red list results, % threatened etc. 
Table 2. in old manuscript
Check methods for tidy tables in R


### 4. Show results of Red List threats

Downlod this using the IDs?
Dig out loop from old red list code

```{r eval = FALSE,echo=TRUE, message = FALSE, warning=FALSE}
#RL_ids = sp_ratings$internalTaxonId

#rlkey = "302e9f949b7c1f3dcc3e1fe64e34e4ee732bdca390f9a09fb76ee8c30ec83124"

#test = rl_threats(id = 30747, key = rlkey)
#test$result
```

### 5. Growth of PAs across Horn of Africa and Aloe ranges

Use WDPA points to generate buffer - use to get max PA area per year later

5a. Use WDPA polys to get minimum PA area per year

```{r eval = FALSE, echo=TRUE, message = FALSE, warning=FALSE}
# read in all the PA points data
PA_polys = st_read("01_data/WDPA_polys_Nov_2018.shp")

# remove those that have no designation status
PA_polys_des = subset(PA_polys, STATUS == "Designated")

# reduce down to something more manageable - filter on countries where points occur
# use ISO code to filter

PA_poly_sub = subset(PA_polys_des, ISO3 == "BFA" |
                       ISO3 == "COD"|
                       ISO3 == "DJI"|
                       ISO3 == "ERI"|
                       ISO3 == "ETH"|
                       ISO3 == "KEN"|
                       ISO3 == "NGA"|
                       ISO3 == "SOM"|
                       ISO3 == "SSD"|
                       ISO3 == "SDN"|
                       ISO3 == "TZA"|
                       ISO3 == "TGO"|
                       ISO3 == "UGA"
                     )
                     
# included points must have a year and an area
includedpol = subset(PA_poly_sub, REP_AREA > "0")
includedpol = subset(includedpol, STATUS_YR > "0")

# now set the projection
included_prj_pol = includedpol %>% st_transform(102022)

# get the unique years to feed into the loop later
uniqueyears = unique(included_prj_pol$STATUS_YR)

# for loop that pulls out each year in dataset

# writes to file
for (i in uniqueyears){
  PA = subset(included_prj_pol,included_prj_pol$STATUS_YR <= i) # query shape year 
  PA_union = st_union(PA, by_feature = FALSE) # dissolves polys
  PA_path = paste0("poly_", i, ".shp")
  write_sf(PA_union, PA_path, driver="ESRI Shapefile")
}
```



### 5b. add points and buffer, then merge with existing PA polys (a) to get maximum PA area 

```{r eval = FALSE, echo=TRUE, message = FALSE, warning=FALSE}
# read in all the PA points data
PA_points = sf::read_sf("01_data/WDPA_points_Nov_2018.shp")

# remove those that have no designation status
PA_points_des = subset(PA_points, STATUS == "Designated")

# reduce down to something more manageable - filter on countries where points occur
# use ISO code to filter

PA_point_sub = subset(PA_points_des, ISO3 == "BFA" |
                       ISO3 == "COD"|
                       ISO3 == "DJI"|
                       ISO3 == "ERI"|
                       ISO3 == "ETH"|
                       ISO3 == "KEN"|
                       ISO3 == "NGA"|
                       ISO3 == "SOM"|
                       ISO3 == "SSD"|
                       ISO3 == "SDN"|
                       ISO3 == "TZA"|
                       ISO3 == "TGO"|
                       ISO3 == "UGA"
                     )

# included points must have a year and an area
includedpoints = subset(PA_point_sub, REP_AREA > "0")
includedpoints = subset(includedpoints, STATUS_YR > "0")

# now set the projection
included_prj_points = includedpoints %>% st_transform(102022)

# get the unique years to feed into the loop later
uniqueyears_points = unique(included_prj_points$STATUS_YR)

# combine the unique years for points and polys (could be different)
years = c(uniqueyears,uniqueyears_points )

# combine points and polys in the same for loop

for (i in years){

  # for polys - query on year and union
  PA = subset(included_prj_pol,included_prj_pol$STATUS_YR <= i) # query poly shape year 
  PA_union = st_union(PA, by_feature = FALSE)
  
  # for points - query on year, then buffer and union
  PA_points = subset(included_prj_points,included_prj_points$STATUS_YR <= i) # query point shape year 
  PA_point_buff = st_buffer(PA_points, dist=sqrt(PA_points$REP_AREA/pi)*1000)
  PA_point_buff_union = st_union(PA_point_buff, by_feature = FALSE)

  # combine features
  comb = st_union(PA_union,  PA_point_buff_union , by_feature = FALSE)

  # save them down
  PA_path = paste0("comb_", i, ".shp")
  write_sf( comb, PA_path, driver="ESRI Shapefile")
}
```

### 6.REPRESENTATION- Using 2km (5 and 10) buffers, chart the number of species that have at least part of range (>= 2km2 protected)

```{r echo=TRUE, message = FALSE, warning=FALSE}

# read in all the occurrence data
occ_sf = sf::st_as_sf(all_occs_rating, coords = c("longitude","latitude"), crs = 4326)

# project to equal area
occ_sf_prj = occ_sf %>% st_transform(102022)

# buffer points 2km
buff_2km = st_buffer(occ_sf_prj, dist=2000)
buff_5km = st_buffer(occ_sf_prj, dist=5000)
buff_10km = st_buffer(occ_sf_prj, dist=10000)
buff_20km = st_buffer(occ_sf_prj, dist=20000)

# save it down for use in Arcmap
write_sf(buff_2km, "buff_2km.shp", driver="ESRI Shapefile")
write_sf(buff_5km, "buff_5km.shp", driver="ESRI Shapefile")
write_sf(buff_10km, "buff_10km.shp", driver="ESRI Shapefile")
write_sf(buff_20km, "buff_20km.shp", driver="ESRI Shapefile")

```

Open in ArcMap
*Dissolve on 'binomial' field so that you have single shape for each species
*Set statistics field as 'RedListCategory' and statistic as 'first' so that you can query this later
*Run intersect tool (default settings) on Minimum PA layer to get PA and Aloe range overlap
*Add area
*Save the output


Now get the count of species per year

```{r echo=TRUE, message = FALSE, warning=FALSE}
### Count of species that have part of range protected (more than 2km?)

# change this to a function

int_area2km = read.csv("04_outputs/min_2km_intersect.csv")
int_area5km = read.csv("04_outputs/min_5km_intersect_out.csv")
int_area10km = read.csv("04_outputs/min_10km_intersect_out.csv")
int_area20km = read.csv("04_outputs/min_20km_intersect_out.csv")

# funciton to take intersect and count unique species
count_sp_buff = function(buff, fname){
  
  # take the intersected data (species with some area protected) and group by species name
  int_area = buff %>% group_by(binomil) 
  
  # count number of species by year
  count_sp = int_area %>% group_by(FID_1) %>% summarise(count = n_distinct(binomil))
  
  # write the file for charting
  path = getwd()
  write.csv(count_sp, paste0(path,"/04_outputs/",fname,".csv"))
  
}

sp_count_2km = count_sp_buff(int_area2km, fname = "2km_buff_sp_count")
sp_count_5km = count_sp_buff(int_area5km, fname = "5km_buff_sp_count")
sp_count_10km = count_sp_buff(int_area10km, fname = "10km_buff_sp_count")
sp_count_20km = count_sp_buff(int_area20km, fname = "20km_buff_sp_count")

# merge these data and plot them up - ggplot line?




##########################################
# now for only the threatened species

thr_int_area2km = int_area2km
thr_int_area5km = int_area5km
thr_int_area10km = int_area10km
thr_int_area20km = int_area20km

count_thr_sp_buff = function(buff, fname){
  
  # query out threatened species only
  threat_int_area = subset(buff, FIRST_rdls == "Critically Endangered"  |
                           FIRST_rdls == "Endangered" |
                           FIRST_rdls == "Vulnerable"
                           )
  
  # take the intersected data (species with some area protected) and group by species name
  int_area = threat_int_area %>% group_by(binomil) 
  
  # count number of species by year
  count_sp = int_area %>% group_by(FID_1) %>% summarise(count = n_distinct(binomil))
  
  # write the file for charting
  path = getwd()
  write.csv(count_sp, paste0(path,"/04_outputs/",fname,".csv"))
  
}

thr_sp_count_2km = count_thr_sp_buff(thr_int_area2km, fname = "2km_buff_thr_sp_count")
thr_sp_count_5km = count_thr_sp_buff(thr_int_area5km, fname = "5km_buff_thr_sp_count")
thr_sp_count_10km = count_thr_sp_buff(thr_int_area10km, fname = "10km_buff_thr_sp_count")
thr_sp_count_20km = count_thr_sp_buff(thr_int_area20km, fname = "20km_buff_thr_sp_count")


```


```{r echo=TRUE, message = FALSE, warning=FALSE}
### 6b Butchart chart. find the % of the range that is protected - then count how many are above/below target thresholds

# combine the year and species name
binom_year = tidyr::unite(int_area2km, binomil, FID_1, col = "comb", remove = FALSE)

# get sum of area by binom/year combination
comb_area = binom_year %>% group_by(comb) %>% summarise(sum = sum(new_area))

# now separate year and binomial
comb_area_yr = tidyr::separate(comb_area, comb, into = c("name", "year"), sep = "_")

# add total range size for each species
tot_area2km = read.csv("04_outputs/2km_buff_sp_area.csv")

# merge protected intersect area with total area to work out %
colnames(tot_area2km)[which(names(tot_area2km) == "binomil")] = "name"
all_area_2km = merge(tot_area2km, comb_area_yr, by = "name")

# now get the proporiton of each species that is protected, for each year
all_area_2km$prop = all_area_2km$sum/all_area_2km$Sum_Area

# subset on species hitting x% (area of range protected) target
all_area_2km_0_2percent = subset(all_area_2km, prop < 0.2)
all_area_2km_0_2percent = all_area_2km_0_2percent %>% group_by(year) %>% summarise(count =  n_distinct(name))

all_area_2km_2_49percent = subset(all_area_2km, prop >= 0.2 & prop < 0.499)
all_area_2km_2_49percent = all_area_2km_2_49percent %>% group_by(year) %>% summarise(count =  n_distinct(name))

all_area_2km_50_100percent = subset(all_area_2km, prop >= 0.50)
all_area_2km_50_100percent = all_area_2km_50_100percent %>% group_by(year) %>% summarise(count =  n_distinct(name))

# save them down
write.csv(all_area_2km_0_2percent, paste0(path,"/04_outputs/all_area_2km_0_2percent.csv"))
write.csv(all_area_2km_2_49percent, paste0(path,"/04_outputs/all_area_2km_2_49percent.csv"))
write.csv(all_area_2km_50_100percent, paste0(path,"/04_outputs/all_area_2km_50_100percent.csv"))

#################### SAME FOR THREATENED SP ################
### 6b Butchart chart. find the % of the range that is protected - then count how many are above/below target thresholds

# query out the threatened sp.
# query out threatened species only
threat_int_area2km = subset(int_area2km, FIRST_rdls == "Critically Endangered"  |
                           FIRST_rdls == "Endangered" |
                           FIRST_rdls == "Vulnerable"
                           )

# combine the year and species name
binom_year = tidyr::unite(threat_int_area2km, binomil, FID_1, col = "comb", remove = FALSE)

# get sum of area by binom/year combination
comb_area = binom_year %>% group_by(comb) %>% summarise(sum = sum(new_area))

# now separate year and binomial
comb_area_yr = tidyr::separate(comb_area, comb, into = c("name", "year"), sep = "_")

# add total range size for each species
tot_area2km = read.csv("04_outputs/2km_buff_sp_area.csv")

# merge protected intersect area with total area to work out %
colnames(tot_area2km)[which(names(tot_area2km) == "binomil")] = "name"
all_area_2km = merge(tot_area2km, comb_area_yr, by = "name")

# now get the proporiton of each species that is protected, for each year
all_area_2km$prop = all_area_2km$sum/all_area_2km$Sum_Area

# subset on species hitting x% (area of range protected) target
all_area_2km_0_2percent_thr = subset(all_area_2km, prop < 0.2)
all_area_2km_0_2percent_thr = all_area_2km_0_2percent_thr %>% group_by(year) %>% summarise(count =  n_distinct(name))

all_area_2km_2_49percent_thr = subset(all_area_2km, prop >= 0.2 & prop < 0.499)
all_area_2km_2_49percent_thr = all_area_2km_2_49percent_thr %>% group_by(year) %>% summarise(count =  n_distinct(name))

all_area_2km_50_100percent_thr = subset(all_area_2km, prop >= 0.50)
all_area_2km_50_100percent_thr = all_area_2km_50_100percent_thr %>% group_by(year) %>% summarise(count =  n_distinct(name))

# save them down
write.csv(all_area_2km_0_2percent_thr, paste0(path,"/04_outputs/all_area_2km_0_2percent_thr.csv"))
write.csv(all_area_2km_2_49percent_thr, paste0(path,"/04_outputs/all_area_2km_2_49percent_thr.csv"))
write.csv(all_area_2km_50_100percent_thr, paste0(path,"/04_outputs/all_area_2km_50_100percent_thr.csv"))
```

### 8. How could PAs be extended in a way that incurs minimal area, but maximum coverage of (threatened) species

```{r echo=TRUE, message = FALSE, warning=FALSE}

points_vs_sites = read.csv("04_outputs/Greedy/species_by_sites.txt")

# filter out species already protected
basic = subset(points_vs_sites, protected != "alreadyPA")


testing = greedy(points_vs_sites)

repeated = (replicate(5, greedy(basic)))
repeat_df = as.data.frame(do.call(rbind,repeated))

sub_repeat = subset(repeat_df, select = c("V2","V3","V4","V5","V6","V7","V8","V9","V10"))

df_freq = names(which.max(table(sub_repeat$V2)))


sites_species = points_vs_sites

# make the function to get sp max and remove from list until you get to 0
# provides a ranking
greedy = function(sites_species){
  
  # some kind of counter to check when you are down to no more rows in the table
  
  testcount = nrow(sites_species)
  
  res.all = data.frame(max = "", area = "", stringsAsFactors = FALSE)
  
  while ((testcount)>'0'){
  # count species per site
  max_sp_site = sites_species %>% 
    group_by(Id) %>% 
    summarise(count =  n_distinct(binomial)) %>% 
    filter(count == max(count)) %>%
    select(Id) %>%
    sample_n(1) # when there is a tie, it adds a random selection, so not every run is the same
  
  max_sp_site = as.character(max_sp_site)
  print(max_sp_site)
  
  # get the list of species at that site
  max_species = subset(sites_species, Id == max_sp_site) %>% group_by(binomial) %>% count(binomial) 
  
  # make a list to feed into the filter
  max_sp_list = as.list(max_species)
  #print(max_sp_list)
  
  # subset these species from the base list
  sp_site_sub = sites_species %>% filter(!binomial %in% max_sp_list$binomial)
  
  # subset to remove site, if not already removed
  new_sites_species = sp_site_sub %>% filter(Id != max_sp_site) # CHECK THIS>>>>
  
  # now get the area for the site
  max_area = subset(sites_species, Id == max_sp_site, select = buff_area)
  max_area = max_area %>% filter(row_number(buff_area)==1) 
  colnames(max_area)[which(names( max_area) == "buff_area")] = "area"
  max_area = as.character(max_area)
  print(max_area)
  
  res = data.frame(max = max_sp_site, area = max_area, stringsAsFactors = FALSE)
  res.all = bind_rows(res.all,res)
  
  testcount = nrow(new_sites_species)
  
  # stop running when area doesn't increase after x runs
  
  }

  return(res.all)
}

```


```{r echo=TRUE, message = FALSE, warning=FALSE}
BGCI.Plants <- read.csv("01_data/BGCI_Plant_Search/BGCI-Plants.csv")

Sp_list <- spec_count

# separate binomial
Sp_list = separate(Sp_list, binomial, c("genus", "Species"), sep = " ")

# now merge
Sp_list_merge = merge(Sp_list, BGCI.Plants, by = "Species")

# tidy up
BGCI_tidy = subset(Sp_list_merge, Genus.Hybrid == "" & 
                     Species.Hybrid == "" &
                     Infraspecific.Rank == "" &
                     Infraspecific.Epithet == "" &
                     Cultivar == "", 
                   select = c(Genus, Species,No..of.ex.situ.sites.worldwide,IUCN.Red.List ))

# group by threatened or not threatened
BGCI_tidy = mutate(BGCI_tidy, Threat.Stat = case_when(IUCN.Red.List == "Least Concern" ~ "Not threatened",
                                                      IUCN.Red.List == "Near Threatened" ~ "Not threatened",
                                                      IUCN.Red.List == "Endangered" ~ "Threatened",
                                                      IUCN.Red.List == "Critically Endangered" ~ "Threatened",
                                                      IUCN.Red.List == "Vulnerable" ~ "Threatened",
                                                      IUCN.Red.List == "Data Deficient" ~ "Data Deficient")) 

# box plot? and stat test
boxplot(No..of.ex.situ.sites.worldwide~Threat.Stat,data=BGCI_tidy, main="threat stat vs bot garden colls", 
  	xlab="threat status", ylab="number of botanical gardens collections per species")


BGCI_tidy_count = BGCI_tidy %>% group_by(Threat.Stat) %>% summarise(n = n())

# write it down
write.csv(Sp_list_merge, paste0("BGCI.csv"))
```
