---
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Methods description for Aloes paper

This document describes methods used in the paper **'Aloes Horn of Africa'**. Each heading relates to sections in the manuscript. See [Aloes_Horn_Diversity](https://github.com/stevenpbachman/Aloes_Horn_Diversity) for raw data and related documents.


### Libraries

```{r, message=FALSE}
library(dplyr)
library(tidyr)
library(readr)
library(ggplot2)
library(rgdal)
library(maptools)
library(sp)
library(rgeos)
library(sf)
library(rredlist)
```

### 1. Histogram of specimen collections 

*Aloe occurrence data*

Specimen data were derived from a database of collections from herbaria including ETH, K, EA, UPS, PRE, and some from FT and other institutions. The taxonomic treatments were verified by one of us (Demissew).
Occurrence data were submitted to the IUCN Red List as supporting information for the global Red List assessments. 

Occurrence data were queried from the [IUCN Red List](https://www.iucnredlist.org/) using the Advanced query and the following filters:

Taxonomy: Genus = Aloe
Land Regions: Country = Djibouti, Eritrea, Ethiopia, Somalia, South Sudan, Sudan 

```{r echo=TRUE, message = FALSE, warning=FALSE}
# import specimens
specimens_all<- read_csv("01_data/Aloe_points_csv.csv")

# add extra range data for two species that didn't have point data
specimens_extra = read_csv("01_data/Aloe_points_extra.csv")

# merge to get full dataset
all_occs = rbind(specimens_all,specimens_extra)

# find unique records (species) and count number of occurrences per species
spec_count = specimens_all %>% group_by(binomial) %>% count()

# plot as histogram and show mean
ggplot(spec_count, aes(x = n)) + 
geom_histogram(color = 'darkgray', fill = 'white', binwidth = 2) +
labs(x = 'Number of occurrence records', y = 'Count of species') +
geom_vline(aes(xintercept = mean(spec_count$n, na.rm = TRUE), show.legend = FALSE)) +
scale_x_continuous(breaks = c(10, 20, 30, 40, 50, 60))

# save down the plot
path = getwd()
ggsave(paste0(path,"/03_figs/fig_1.png"), width = 10)

# calculate the mean number of occurrences per species
spec_mean = mean(spec_count$n)

```

### 2. Study area – show all points, and show study region countries where species are from.

Done in arcmap - explain

### 3. Aloe Red List assessment data - summarise categories and threat data:

Red List assessment data were queried from the [IUCN Red List](https://www.iucnredlist.org/) using the Advanced query and the following filters:

Taxonomy: Genus = Aloe
Land Regions: Country = Djibouti, Eritrea, Ethiopia, Somalia, South Sudan, Sudan 

The following tables were downloaded:
assessments: Species ID and Red List category
taxonomy: Species ID and taxonomic information (linked by species ID)

```{r echo=TRUE, message = FALSE, warning=FALSE}
# import data
assessments = read_csv("01_data/IUCN_downloads/assessments.csv")
taxonomy = read_csv("01_data/IUCN_downloads/taxonomy.csv")

# get single file with names and ratings
sp_ratings = merge(assessments, taxonomy, by = "internalTaxonId")

# update points with Red List Rating
colnames(all_occs)[which(names(all_occs) == "id_no")] = "internalTaxonId"
all_occs_rating = merge(all_occs, sp_ratings, by = "internalTaxonId")
```

*Summary of final Red List assessment ratings*

Replicate method to calculate table of red list results, % threatened etc. 
Table 2. in old manuscript
Check methods for tidy tables in R


### 4. Show results of Red List threats

Downlod this using the IDs?
Dig out loop from old red list code

```{r eval = FALSE,echo=TRUE, message = FALSE, warning=FALSE}
#RL_ids = sp_ratings$internalTaxonId

#rlkey = "302e9f949b7c1f3dcc3e1fe64e34e4ee732bdca390f9a09fb76ee8c30ec83124"

#test = rl_threats(id = 30747, key = rlkey)
#test$result
```

### 5. Growth of PAs across Horn of Africa and Aloe ranges

Use WDPA points to generate buffer - use to get max PA area per year later

5a. Use WDPA polys to get minimum PA area per year

```{r eval = FALSE, echo=TRUE, message = FALSE, warning=FALSE}
# read in all the PA points data
PA_polys = st_read("01_data/WDPA_polys_Nov_2018.shp")

# remove those that have no designation status
PA_polys_des = subset(PA_polys, STATUS == "Designated")

# reduce down to something more manageable - filter on countries where points occur
# use ISO code to filter

PA_poly_sub = subset(PA_polys_des, ISO3 == "BFA" |
                       ISO3 == "COD"|
                       ISO3 == "DJI"|
                       ISO3 == "ERI"|
                       ISO3 == "ETH"|
                       ISO3 == "KEN"|
                       ISO3 == "NGA"|
                       ISO3 == "SOM"|
                       ISO3 == "SSD"|
                       ISO3 == "SDN"|
                       ISO3 == "TZA"|
                       ISO3 == "TGO"|
                       ISO3 == "UGA"
                     )
                     
# included points must have a year and an area
includedpol = subset(PA_poly_sub, REP_AREA > "0")
includedpol = subset(includedpol, STATUS_YR > "0")

# now set the projection
included_prj_pol = includedpol %>% st_transform(102022)

# get the unique years to feed into the loop later
uniqueyears = unique(included_prj_pol$STATUS_YR)

# for loop that pulls out each year in dataset

# writes to file
for (i in uniqueyears){
  PA = subset(included_prj_pol,included_prj_pol$STATUS_YR <= i) # query shape year 
  PA_union = st_union(PA, by_feature = FALSE) # dissolves polys
  PA_path = paste0("poly_", i, ".shp")
  write_sf(PA_union, PA_path, driver="ESRI Shapefile")
}
```



### 5b. add points and buffer, then merge with existing PA polys (a) to get maximum PA area 

```{r eval = FALSE, echo=TRUE, message = FALSE, warning=FALSE}
# read in all the PA points data
PA_points = sf::read_sf("01_data/WDPA_points_Nov_2018.shp")

# remove those that have no designation status
PA_points_des = subset(PA_points, STATUS == "Designated")

# reduce down to something more manageable - filter on countries where points occur
# use ISO code to filter

PA_point_sub = subset(PA_points_des, ISO3 == "BFA" |
                       ISO3 == "COD"|
                       ISO3 == "DJI"|
                       ISO3 == "ERI"|
                       ISO3 == "ETH"|
                       ISO3 == "KEN"|
                       ISO3 == "NGA"|
                       ISO3 == "SOM"|
                       ISO3 == "SSD"|
                       ISO3 == "SDN"|
                       ISO3 == "TZA"|
                       ISO3 == "TGO"|
                       ISO3 == "UGA"
                     )

# included points must have a year and an area
includedpoints = subset(PA_point_sub, REP_AREA > "0")
includedpoints = subset(includedpoints, STATUS_YR > "0")

# now set the projection
included_prj_points = includedpoints %>% st_transform(102022)

# get the unique years to feed into the loop later
uniqueyears_points = unique(included_prj_points$STATUS_YR)

# combine the unique years for points and polys (could be different)
years = c(uniqueyears,uniqueyears_points )

# combine points and polys in the same for loop

for (i in years){

  # for polys - query on year and union
  PA = subset(included_prj_pol,included_prj_pol$STATUS_YR <= i) # query poly shape year 
  PA_union = st_union(PA, by_feature = FALSE)
  
  # for points - query on year, then buffer and union
  PA_points = subset(included_prj_points,included_prj_points$STATUS_YR <= i) # query point shape year 
  PA_point_buff = st_buffer(PA_points, dist=sqrt(PA_points$REP_AREA/pi)*1000)
  PA_point_buff_union = st_union(PA_point_buff, by_feature = FALSE)

  # combine features
  comb = st_union(PA_union,  PA_point_buff_union , by_feature = FALSE)

  # save them down
  PA_path = paste0("comb_", i, ".shp")
  write_sf( comb, PA_path, driver="ESRI Shapefile")
}
```

### 6.REPRESENTATION- Using 2km (5 and 10) buffers, chart the number of species that have at least part of range (>= 2km2 protected)

```{r echo=TRUE, message = FALSE, warning=FALSE}

# read in all the occurrence data
occ_sf = sf::st_as_sf(all_occs_rating, coords = c("longitude","latitude"), crs = 4326)

# project to equal area
occ_sf_prj = occ_sf %>% st_transform(102022)

# buffer points 2km
buff_2km = st_buffer(occ_sf_prj, dist=2000)

# save it down for use in Arcmap
PA_path = paste0("buff_2km", ".shp")
write_sf(buff_2km, PA_path, driver="ESRI Shapefile")

```

Open in ArcMap
Dissolve on 'binomial' field so that you have single shape for each species
Set statistics field as 'RedListCategory' and statistic as 'first' so that you can query this later
Run intersect tool (default settings) on Minimum PA layer to get PA and Aloe range overlap


Now get the count of species per year

```{r echo=TRUE, message = FALSE, warning=FALSE}
### Count of species that have part of range protected (more than 2km?)

# take the intersected data (species with some area protected) and group by species name
int_area = min_2km_intersect_out %>% group_by(binomial) 

# count number of species by year
count_sp = int_area %>% group_by(FID_1) %>% summarise(count = n_distinct(binomial))

# write the file for charting
path = getwd()
write.csv(count_sp, paste0(path,"/04_outputs/sp_per_year_min_2km.csv"))
```


